---
title: "Apoe Mouse Project"
author: "Christine Snow"
date: "2026-06-18"
output: html_document
---

Download the libraries needed.
```{r libraries}
library(ggtext) #enables complex formatted plot labels
library(tidyverse) #core
library(ggrepel) #repel overlapping text labels in ggplot
library(ggplot2) #visualization
library(readxl) #read xl files and turn into dataframes
library(DESeq2) #identifies differentially expressed genes or transcripts in RNA seq
library(edgeR) #Differential expression analysis of sequence count data
library(ComplexHeatmap) # arrange multiple heatmaps and supports various annotation graphics
library(circlize) #efficient way for the visualization of large amounts of information
library(fgsea) #algorithm for fast gene set enrichment analysis
library(data.table) # designed for rapid data manipulation and aggregation
library(patchwork) #makes it simple to combine separate ggplots into the same graphic
library(devEMF) # EMF (for Illustrator)
library(grid)
```

Version numbers for packages used
```{r}
sapply(c("ggtext", "tidyverse", "ggrepel", "ggplot2", "readxl", "DESeq2", "edgeR", "ComplexHeatmap", "circlize", "fgsea", "data.table", "patchwork", "devEMF", "grid"), function(pkg) as.character(packageVersion(pkg)))
R.version.string
```

Within 'raw_data' folder there are: genotypes_combined.csv, gene_counts_salmon_2nd_analysis.csv, sample_list.xlsx

```{r read files}
#load the three input files
combined <- read.csv('raw_data/genotypes_combined.csv')
counts   <- read.csv('raw_data/gene_counts_salmon_2nd_analysis.csv')
samples  <- read_excel("raw_data/sample_list.xlsx")

#Gene Ids look like: ENSMUSG00000085882.1 --> Need to remove the .x at the end to make sure its interchangeable. 
#Also, the row names need to be the samples.

combined$gene_id <- sub("\\..*", "", combined$gene_id) 
counts$geneID    <- sub("\\..*", "", counts$geneID)

coldata <- column_to_rownames(samples, var = "Sample Submission Name") 

gene_names <- combined %>%
  select(gene_id, gene_name, gene_type) %>%
  distinct()
```

Round Salmon's fractional counts to whole numbers, then turn it into a matrix
```{r counts}
counts_for_matrix <- counts
if ("geneID" %in% colnames(counts_for_matrix)) {
  counts_for_matrix <- counts_for_matrix %>% column_to_rownames("geneID")
}

cts <- counts_for_matrix %>%
  round() %>%
  as.matrix()

# Every sample in coldata should match a column in the counts.
all(rownames(coldata) %in% colnames(cts)) # this should print TRUE — if FALSE, the sample names don't line up
```

Create two new columns to cleanly group the genes by apoe or splice effect
```{r coldata}
coldata <- coldata %>%
  mutate(
    apoe = case_when(
      grepl("APOE3", Genotype) ~ "E3",
      grepl("APOE4", Genotype) ~ "E4"
    ),
    splicing = case_when(
      grepl("\\+ex19", Genotype) ~ "plus",
      grepl("Δex19",  Genotype)  ~ "delta"
    )
  )

#overrides the default and explicitly says which group is the baseline

coldata$apoe <- relevel(factor(coldata$apoe), ref = "E3") 
coldata$splicing <- relevel(factor(coldata$splicing), ref = "plus")
```

Run edgeR

- The main components of an DGEList object are a matrix counts containing the integer counts,
a data.frame samples containing information about the samples or libraries

```{r edgeR}
design <- model.matrix(~ apoe + splicing + apoe:splicing, data = coldata) 
dge <- DGEList(counts = cts, samples = coldata) 

# filter out low-expressed genes 
keep <- filterByExpr(dge, design = design)
dge  <- dge[keep, , keep.lib.sizes = FALSE]
nrow(dge)   # should be ~18,764

# TMM normalisation (edgeR default)
dge <- calcNormFactors(dge, method = "TMM")

# estimate dispersions (robust = TRUE down-weights outlier genes)
dge <- estimateDisp(dge, design, robust = TRUE)

# fit quasi-likelihood model
fit <- glmQLFit(dge, design, robust = TRUE) #adds an extra layer of uncertainty to account for only 3 rep per group

# Normalised CPM counts (used later for heatmap and PCA)
# equivalent to counts(dds, normalized = TRUE) in DESeq2
normalized_counts <- cpm(dge, normalized.lib.sizes = TRUE)
saveRDS(normalized_counts, "edgeR/results/normalized_counts.rds")
```

Extract the three results tables
Column names in edgeR: logFC, logCPM, F, PValue, FDR.
glmQLFTest() tests the coefficient named by `coef`
Summarize the results

```{r}
res_apoe_main_effect <- glmQLFTest(fit, coef = "apoeE4")
res_splicing_main_effect <- glmQLFTest(fit, coef = "splicingdelta")
res_interaction <- glmQLFTest(fit, coef = "apoeE4:splicingdelta")

# decideTests() applies FDR + optional logFC thresholds and prints up/down counts
summary(decideTests(res_apoe_main_effect))
summary(decideTests(res_splicing_main_effect))
summary(decideTests(res_interaction))
```

Convert the results to data frames.
toptags is an edger function to extract data frame from edger results that would otherwise just be raw values
add gene names in by joining to the gene_name table 

```{r convert to dataframes}
df_res_apoe_main_effect <- topTags(res_apoe_main_effect,
                                    n = Inf, sort.by = "none")$table %>% #n = Inf, sort.by = "none" --> all genes in their original order, not just the top ones
  rownames_to_column("gene_id") %>%
  left_join(gene_names, by = "gene_id")

df_res_splicing_main_effect <- topTags(res_splicing_main_effect,
                                        n = Inf, sort.by = "none")$table %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_names, by = "gene_id")

df_res_interaction <- topTags(res_interaction,
                               n = Inf, sort.by = "none")$table %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_names, by = "gene_id")

write_csv(df_res_interaction, file = "deliverables/interaction_effect_df.csv")
```

Filter for significant genes with FDR < .05

```{r filter genes}
filter_df <- function(df) {
  df %>% 
    dplyr::filter(FDR < 0.05)
}

filtered_apoe_main_effect <- filter_df(df_res_apoe_main_effect)
filtered_splicing_main_effect <- filter_df(df_res_splicing_main_effect) 
filtered_interaction <- filter_df(df_res_interaction)
```

Create a master list of genes by combining each gene, deduplicating them, and creating columns indicating which term(s) each gene came from 

```{r master list}
master_list <-
  bind_rows(filtered_apoe_main_effect, filtered_interaction, filtered_splicing_main_effect) %>% #combine the lists
  distinct(gene_id, .keep_all = TRUE) %>% #keep the first time that gene is seen, drop the rest
  select(gene_id, gene_name) %>% # select from that list only the gene name and id
  mutate( #create three columns
    in_apoe= gene_id %in% filtered_apoe_main_effect$gene_id,#if that specific gene_id is in that list, then a boolean expression indicating that
    in_splice = gene_id %in% filtered_splicing_main_effect$gene_id,
    in_interaction = gene_id %in% filtered_interaction$gene_id
  )

# genes that only appear in interaction effect (not significant in apoe or ex19 effect) 

novel_interaction_genes <- master_list %>%
  filter(in_interaction == TRUE &
         in_apoe == FALSE &
         in_splice == FALSE)

write_csv(novel_interaction_genes, file = "extras/novel_interaction_genes.csv")
```


```{r save files}
# saved in RNA_seq_project/filtered_results folder 
write_csv(filtered_apoe_main_effect, file = "filtered_results/filtered_apoe_main_effect.csv")
write_csv(filtered_splicing_main_effect, file = "filtered_results/filtered_splicing_main_effect.csv")
write_csv(filtered_interaction, file = "filtered_results/filtered_interaction.csv")

#in deliverables folder
write_csv(master_list, file = "deliverables/master_gene_list.csv")
```

 
Creating a volcano plot function:
```{r}
volcano_plotting <- function(df, title, x_breaks) {
  df$diffexpressed <- "NO"
  df$diffexpressed[df$FDR < 0.05 & df$logFC > 0] <- "UP"
  df$diffexpressed[df$FDR < 0.05 & df$logFC < 0] <- "DOWN"
  df$delabel <- ifelse(
  df$gene_name %in%
    head(df[order(df$FDR), "gene_name"], 10),
  df$gene_name, NA)
  
  
  ggplot(df,aes(x = logFC, y = -log10(FDR), col = diffexpressed, label = delabel)) +
  geom_vline(xintercept = 0, col = "darkgrey", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), col = "darkgrey", linetype = "dashed", linewidth = 0.5) +
  geom_point(size = 1.5, alpha = 0.6) +
  scale_color_manual(values = c("DOWN" = "steelblue", "NO" = "grey80", "UP" = "darkorange"),
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  labs(color = NULL,
       x = expression("Log"[2]*" (Fold Change)"),
       y = expression("-Log"[10]*" (FDR)")) +
  scale_x_continuous(breaks = x_breaks)  +
  ggtitle(title) +
  geom_text_repel(max.overlaps = Inf, size = 3, color = "black", family = "Helvetica") +
  theme_classic() +
  theme(
    text = element_text(family = "Helvetica"),
    legend.position = "none",
    plot.title = element_text(hjust = 0, face = "bold")
  )
}
```

```{r volcano plot 1}
# --- figure_1: ApoE main effect (E4 vs E3) ---
volcano_plotting(df_res_apoe_main_effect, "APOE4 vs APOE3",seq(-10, 10, 2) )
```

```{r volcano plot 2}
# --- figure_2: splicing main effect (delta vs +ex19) ---
volcano_plotting(df_res_splicing_main_effect, "Apoer2 \u0394ex19 vs. [+ex19]", seq(-5, 5, 2))
```
Publication ready volcano plot:
```{r interaction volcano plot}
# interaction volcano plot

plot_df_interaction <- df_res_interaction 

plot_df_interaction$diffexpressed <- "NO"
plot_df_interaction$diffexpressed[plot_df_interaction$FDR < 0.1 & plot_df_interaction$logFC > 1] <- "UP"
plot_df_interaction$diffexpressed[plot_df_interaction$FDR < 0.1 & plot_df_interaction$logFC < -1] <- "DOWN"
#For FDR it should be < 0.1 and FC should be >1 or <-1
plot_df_interaction$delabel <- ifelse(
  plot_df_interaction$FDR < 0.1 &
  abs(plot_df_interaction$logFC) > 1,
  plot_df_interaction$gene_name,
  NA
)

interaction_effect_volcano_plot <- ggplot(plot_df_interaction,
                   aes(x = logFC, y = -log10(FDR),
                       col = diffexpressed, label = delabel)) +
  geom_vline(xintercept = 1, linetype = "longdash", colour = "#f99d41") +
  geom_vline(xintercept = -1, linetype = "longdash", colour = "#6495ed") +
  geom_hline(yintercept = 1, linetype = "longdash", colour = "grey80") +
  geom_point(size = 1.5, alpha = 1, stroke = 0.5) +
  scale_color_manual(values = c("DOWN" = "steelblue", "NO" = "grey80", "UP" = "darkorange"),
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  labs(color = NULL,
       x = expression("Log"[2]*" (Fold Change)"),
       y = expression("-Log"[10]*" (FDR)")) +
  scale_x_continuous(breaks = seq(-10, 10, 5)) +
  scale_y_continuous(breaks = seq(0, 8, 2))+
  annotate("label",
           x = -Inf, y = Inf,
           label = "ApoE \u00d7 Ex19 interaction",
           hjust = -.05, vjust = 1.3,
           size = 8 / .pt,
           family = "Arial",
           fontface = "bold",
           fill = "white",
           label.size = 0.3) +
  geom_text_repel(max.overlaps = Inf, size = 7 / .pt, color = "black",
                  family = "Arial", box.padding = 0.3, segment.size = 0.4, min.segment.length = .1) +
  theme_bw() +
  theme(
    text = element_text(family = "Arial"),
    legend.position = "none",
    axis.title = element_text(size = 10, family = "Arial"),
    axis.text = element_text(size = 8, family = "Arial"),
    axis.line = element_line(linewidth = 0.75 / 2.13),
    axis.ticks = element_line(linewidth = 0.75 / 2.13)
  )

interaction_effect_volcano_plot

# PDF (visual reference)

ggsave("deliverables/Fig2_PanelA_interaction_volcano.pdf",
       interaction_effect_volcano_plot,
       width = 5, height = 3.5, device = cairo_pdf)

# EMF
ggsave(
  "deliverables/Fig2_PanelA_interaction_volcano.svg",
  interaction_effect_volcano_plot,
  width = 5,
  height = 3.5
)

```


Heatmap of significant genes (Z-scored, grouped by genotype)
rows = all significant genes (from master_list), columns = the 12 samples
split into the 4 genotype groups. Z-score per gene (scale across samples).
normalized_counts here is CPM with TMM-adjusted library sizes.

```{r heatmap all genes} 

col_fun <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#E8A838"))

bottom_ann <- HeatmapAnnotation(
  Group = anno_block(
    gp = gpar(fill = c("#8B9DC3", "#AED6F1", "#5D8A5E", "#A8D5A2")),
    labels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"),
    labels_gp = gpar(fontsize = 10, col = "black")
  ),
  which = "column",
  show_annotation_name = FALSE
)

col_split <- factor(c(rep("+ex19/E3", 3), rep("\u0394ex19/E3", 3),
                      rep("+ex19/E4", 3), rep("\u0394ex19/E4", 3)),
                    levels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"))


sig_genes <- master_list$gene_id
heatmap_mat <- normalized_counts[rownames(normalized_counts) %in% sig_genes, ]
heatmap_mat_scaled <- t(scale(t(heatmap_mat)))

ht <- Heatmap(heatmap_mat_scaled,
              col = col_fun,
              name = "Z-Score",
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              cluster_columns = FALSE,
              column_split = col_split,
              bottom_annotation = bottom_ann,
              column_title = NULL,
              border = FALSE,
              row_dend_side = "left",
              heatmap_legend_param = list(
                title = "Z-Score",
                at = c(-2, -1, 0, 1, 2),
                labels = c("-2", "-1", "0", "1", "2")
              ))
draw(ht)
```


PCA coloured by genotype group
Uses log2 CPM on TMM-normalised counts 

```{r pca}
lcpm <- cpm(dge, log = TRUE)

pca <- prcomp(t(lcpm), scale. = FALSE)
pca_data <- as.data.frame(pca$x[, 1:2]) %>%
  rownames_to_column("sample")

percent_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2))

# attach genotype metadata
pca_data <- pca_data %>%
  left_join(
    coldata %>% rownames_to_column("sample") %>% select(sample, apoe, splicing),
    by = "sample"
  )

pca_data$group <- paste0(ifelse(pca_data$splicing == "plus", "+ex19", "\u0394ex19"),
                          "/", pca_data$apoe)
pca_data$group <- factor(pca_data$group,
                          levels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"))

pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2,
                                  color = group, shape = group, label = sample)) +
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0.15,
               level = 0.95, show.legend = FALSE) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, show.legend = FALSE) +
  scale_color_manual(values = c("+ex19/E3" = "#3B5998", "\u0394ex19/E3" = "#4BACC6",
                                "+ex19/E4" = "#1F5C1F", "\u0394ex19/E4" = "#92C46A")) +
  scale_fill_manual(values  = c("+ex19/E3" = "#3B5998", "\u0394ex19/E3" = "#4BACC6",
                                "+ex19/E4" = "#1F5C1F", "\u0394ex19/E4" = "#92C46A")) +
  scale_shape_manual(values = c("+ex19/E3" = 16, "\u0394ex19/E3" = 18,
                                "+ex19/E4" = 15, "\u0394ex19/E4" = 16)) +
  labs(x = paste0("PC1 (", percent_var[1], "%)"),
       y = paste0("PC2 (", percent_var[2], "%)"),
       color = "Group", shape = "Group") +
  theme_bw() +
  theme(panel.grid.major = element_line(color = "grey90"),
        panel.grid.minor = element_blank(),
        legend.title      = element_text(face = "bold"),
        legend.background = element_rect(color = "grey70"))
pca_data
pca_plot

ggsave("deliverables/pca_plot.svg",
       pca_plot, width = 8, height = 5)
write_csv(pca_data, "deliverables/pca_plot.csv")
```
QC test when switching from deseq to edger

```{r deseq vs edgeR, eval = FALSE}

deseq <- read_csv("archive/10_deseq_results/filtered_interaction.csv")
edger <- read_csv("edgeR/results/filtered_interaction.csv")

deseq <- read_csv("archive/10_deseq_results/filtered_interaction.csv")
edger <- read_csv("edgeR/results/filtered_interaction.csv")

edger_clean <- edger %>%
  dplyr::select(gene_id, gene_name, logFC, FDR) %>%
  dplyr::rename(logFC_edger = logFC, FDR_edger = FDR)

deseq_clean <- deseq %>%
  dplyr::select(gene_id, gene_name, log2FoldChange, padj) %>%
  dplyr::rename(logFC_deseq2 = log2FoldChange, FDR_deseq2 = padj)
# full join keeps all genes from both
comparison <- full_join(deseq_clean, edger_clean, by = c("gene_id", "gene_name")) %>%
  mutate(
    found_in = case_when(
      !is.na(logFC_deseq2) & !is.na(logFC_edger) ~ "Both",
      !is.na(logFC_deseq2) & is.na(logFC_edger)  ~ "DESeq2 only",
      is.na(logFC_deseq2)  & !is.na(logFC_edger) ~ "edgeR only"
    )
  ) %>%
  arrange(found_in, gene_name)

# check counts
table(comparison$found_in)

comparison
# save
 #write_csv(comparison, "edgeR/tables/interaction/DESeq2_vs_edgeR_interaction_comparison.csv")
```

QC check
Dnah14 outlier check — CG08 is not a global outlier, difference < 0.3 CPM

```{r dnah check, eval=FALSE}
dnah14_id <- "ENSMUSG00000047369"

# get normalized counts for Dnah14
dnah14_counts <- norm[dnah14_id, ]
print(dnah14_counts)
```


---Pathway Analysis----

1. Load downloaded pathways in raw_data/ folder for fgsea
```{r load go-pathways}
# 1 GO pathway gene sets
bp_pathways <- fgsea::gmtPathways("raw_data/m5.go.bp.v2025.1.Mm.symbols.gmt")
cc_pathways <- fgsea::gmtPathways("raw_data/m5.go.cc.v2025.1.Mm.symbols.gmt")
mf_pathways <- fgsea::gmtPathways("raw_data/m5.go.mf.v2025.1.Mm.symbols.gmt")

saveRDS(bp_pathways, "raw_data/pathways/bp_pathways.rds")
saveRDS(cc_pathways, "raw_data/pathways/cc_pathways.rds")
saveRDS(mf_pathways, "raw_data/pathways/mf_pathways.rds")

# combined for full-genome runs (sections 5 and 6)
pathways_all <- c(bp_pathways, cc_pathways, mf_pathways)

# per-category list for the loop in section 4
paths <- list(GO_MF = mf_pathways,
              GO_CC = cc_pathways,
              GO_BP = bp_pathways)
```

2. GSEA functions 

```{r gsea-functions}
#build a barplot from a combined top/bottom NES table
make_barplot <- function(combined_nes, title, text_size = 12) {
  combined_nes %>%
    mutate(
      direction = ifelse(NES > 0, "Positive Enrichment", "Negative Enrichment"),
      pathway_clean = pathway %>%
        str_replace_all("_", " ") %>%
        str_to_sentence()
    ) %>%
    ggplot(aes(x = reorder(pathway_clean, NES), y = NES, fill = direction)) +
    geom_col(color = "black", linewidth = 0.5) +
    coord_flip() +
    labs(x = "Pathway",
         y = "Normalized Enrichment Score (NES)",
         fill = "Pathway Enrichment",
         title = title) +
    theme_bw() +
    theme(
      axis.text.y  = element_text(size = text_size),
      axis.text.x  = element_text(size = text_size - 2),
      axis.title   = element_text(size = text_size + 1),
      plot.title   = element_text(size = text_size + 3, face = "bold"),
      legend.text  = element_text(size = text_size)
    ) +
    scale_fill_manual(values = c("Positive Enrichment" = "#FFC541",
                                 "Negative Enrichment" = "#6495ED"))
}

#F-statistic-ranked GSEA 
# F is always positive in edgeR, so sign is restored via logFC direction.
run_gsea_f <- function(res_df, pathway, title) {
  ranked_genes <- res_df %>%
    dplyr::filter(!is.na(gene_name), !is.na(F)) %>%
    group_by(gene_name) %>%
    slice_max(abs(F), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(F_signed = F * sign(logFC)) %>%
    { setNames(.$F_signed, .$gene_name) }

  fgseaRes <- fgsea(pathways = pathway,
                    stats    = ranked_genes,
                    minSize  = 15,
                    maxSize  = 500)

  gsea_df    <- as.data.frame(fgseaRes)
  top_ten    <- head(arrange(gsea_df, desc(NES)), 10)
  bottom_ten <- tail(arrange(gsea_df, desc(NES)), 10)
  combined   <- bind_rows(top_ten, bottom_ten)

  list(plot = make_barplot(combined, title),
       data = combined,
       full_results = fgseaRes)
}
```

3. Run GSEA across all three contrasts × GO categories 

```{r run-gsea-loop}
terms <- list(
  ApoE = df_res_apoe_main_effect,
  Ex19 = df_res_splicing_main_effect,
  interaction = df_res_interaction
)
set.seed(42)
for (term_name in names(terms)) {
  for (cat_name in names(paths)) {

    result <- run_gsea_f(
      terms[[term_name]],
      paths[[cat_name]],
      paste(term_name, cat_name, "F-ranked")
    )

    fig_dir   <- file.path("edgeR/figures/GSEA", term_name)
    table_dir <- file.path("edgeR/tables/GSEA", term_name)
    dir.create(fig_dir,   recursive = TRUE, showWarnings = FALSE)
    dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

    base <- paste0("GSEA_", term_name, "_", cat_name)

    ggsave(file.path(fig_dir, paste0(base, "_top10_posneg.pdf")),
           result$plot, width = 9, height = 6)
    ggsave(file.path(fig_dir, paste0(base, "_top10_posneg.png")),
           result$plot, width = 9, height = 6, dpi = 300)

    result$full_results %>%
      mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";")) %>%
      write_csv(file.path(table_dir, paste0(base, ".csv")))
  }
}

```


4. Pathway analysis
run against pathways_all (BP + CC + MF combined).

```{r interaction-comparison}
set.seed(42)
result_f <- fgsea(
  pathways = pathways_all,
  stats    = {
    df_res_interaction %>%
      dplyr::filter(!is.na(gene_name), !is.na(F)) %>%
      group_by(gene_name) %>%
      slice_max(abs(F), n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      mutate(F_signed = F * sign(logFC)) %>%
      { setNames(.$F_signed, .$gene_name) }
  },
  minSize = 15, maxSize = 500,
  nPermSimple = 100000
) %>%
  mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ";"))

arrange(result_f, padj)
#arrange(result_logfc, padj)
```

```{r gsea bar chart}
#GSEA bar chart function 
make_gsea_bar <- function(df, title, n_pos = 10, n_neg = 10) {
  
  top_pos <- df %>% arrange(desc(NES)) %>% head(n_pos)
  top_neg <- df %>% arrange(NES)       %>% head(n_neg)
  
  combined <- bind_rows(top_pos, top_neg) %>%
    distinct(pathway, .keep_all = TRUE) %>%
    mutate(
      direction = ifelse(NES > 0, "Positive", "Negative"),
      pathway_clean = pathway %>%
        str_remove("^GOBP_|^GOCC_|^GOMF_") %>%
        str_replace_all("_", " ") %>%
        str_to_sentence() #%>%
        #str_wrap(width = 40)
    )
  
  #  corrections for pathway labels
  label_fixes <- c(
    "9plus2 motile cilium" = "9+2 motile cilium",
    "Metallo sulfur cluster assembly" = "Metallo-sulfur cluster assembly",
    "Nadh dehydrogenase complex" = "NADH dehydrogenase complex",
    "Pcg protein complex" = "PcG protein complex"
  )
  
  for (old in names(label_fixes)) {
    combined$pathway_clean[combined$pathway_clean == old] <- label_fixes[old]
  }
  
  combined$pathway_clean <- str_replace(
    combined$pathway_clean,
    "(?i)rna polymerase ii transcription\\s*\\n?regulator complex",
    "RNA polymerase II transcription\nregulator complex"
  )
  
  ggplot(combined, aes(x = NES,
                       y = reorder(pathway_clean, NES),
                       fill = direction)) +
    geom_col(color = "black", linewidth = 0.5) +
    geom_text(data = . %>% dplyr::filter(padj < 0.05 & NES > 0),
              aes(label = "*"),
              hjust = 1.3, vjust = 0.7,
              size = 6, color = "black") +
    geom_text(data = . %>% dplyr::filter(padj < 0.05 & NES < 0),
              aes(label = "*"),
              hjust = -0.3, vjust = 0.7,
              size = 6, color = "black") +
    geom_vline(xintercept = 0, linewidth = 0.3, color = "black") +
    scale_fill_manual(values = c("Positive" = "#FFC642",
                                 "Negative" = "#6E91CB")) +
    labs(title = title, x = "NES", y = NULL) +
    theme_bw() +
    theme(
      text = element_text(family = "Helvetica"),
      panel.border = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_text(size = 7),
      axis.text.y = element_text(size = 9),
      axis.text.x = element_text(size = 8),
      plot.title = element_text(size = 10, face = "bold"),
      legend.position = "none"
    )
}

# Load GSEA results 
# F-statistic ranked
gsea_f <- read_csv("edgeR/tables/GSEA/interaction/GSEA_interaction_rank_F.csv")


# Split by GO category  
bp_f <- gsea_f %>% dplyr::filter(str_detect(pathway, "^GOBP_"))
cc_f <- gsea_f %>% dplyr::filter(str_detect(pathway, "^GOCC_"))
mf_f <- gsea_f %>% dplyr::filter(str_detect(pathway, "^GOMF_"))
#bp_logfc <- gsea_logfc %>% dplyr::filter(str_detect(pathway, "^GOBP_"))
#cc_logfc <- gsea_logfc %>% dplyr::filter(str_detect(pathway, "^GOCC_"))

#F-ranked panels 
panel_b_bp_f <- make_gsea_bar(bp_f, "GO Biological Process", n_pos = 10, n_neg = 10)
panel_b_cc_f <- make_gsea_bar(cc_f, "GO Cellular Component", n_pos = 10, n_neg = 10)
panel_b_mf_f <- make_gsea_bar(mf_f, "GO Molecular Function", n_pos = 10, n_neg = 10)
panel_b_bp_f
panel_b_cc_f
panel_b_mf_f


  
ggsave("deliverables/Fig2_PanelB_GSEA_BP.pdf",
       panel_b_bp_f, width = 8, height = 5, device = cairo_pdf)
ggsave("deliverables/Fig2_PanelB_GSEA_CC.pdf",
       panel_b_cc_f, width = 5, height = 3.5, device = cairo_pdf)
ggsave("deliverables/Fig2_PanelB_GSEA_MF.pdf",
       panel_b_mf_f, width = 8, height = 3.5, device = cairo_pdf)

ggsave("deliverables/Fig2_PanelB_GSEA_BP.svg",
       panel_b_bp_f, width = 8, height = 5)
ggsave("deliverables/Fig2_PanelB_GSEA_CC.svg",
       panel_b_cc_f, width = 5, height = 3.5)
ggsave("deliverables/Fig2_PanelB_GSEA_MF.svg",
       panel_b_mf_f, width = 8, height = 3.5)



```
Pull specific pathways 

```{r}
gsea_f %>% 
  dplyr::filter(pathway %in% c(
    "GOCC_9PLUS2_MOTILE_CILIUM",
    "GOCC_MOTILE_CILIUM",
    "GOBP_CILIUM_MOVEMENT",
    "GOBP_SODIUM_ION_EXPORT_ACROSS_PLASMA_MEMBRANE",
    "GOBP_CILIUM_OR_FLAGELLUM_DEPENDENT_CELL_MOTILITY"
  )) %>%
  dplyr::select(pathway, padj, NES)
```

Find the top three genes in apoe effect and exon19 to add to the heatmap

```{r}
top3_apoe <- filtered_apoe_main_effect %>% # finding the top 3 genes arranged by FDR in the apoe effect list
  dplyr::arrange(FDR) %>%
  dplyr::slice_head(n = 3) %>%
  dplyr::pull(gene_name)


top3_ex19 <- filtered_splicing_main_effect %>% #same as above but with exon19 effect
  dplyr::arrange(FDR) %>%
  dplyr::slice_head(n = 3) %>%
  dplyr::pull(gene_name)
top3_ex19
```

```{r save-interaction-deliverables}
dir.create("edgeR/tables/GSEA/interaction", recursive = TRUE, showWarnings = FALSE)

write_csv(result_f, "edgeR/tables/GSEA/interaction/GSEA_interaction_rank_F.csv")
```


5. Cilia leading-edge heatmap 
pool GOBP_CILIUM_MOVEMENT leading-edge genes across all three contrasts, rank by FDR 

```{r cilia-leading-edge}
# creating a function that searches for this pathway specifically "GOBP CILIUM MOVEMENT" 

get_leading_edge <- function(df) {
  df %>%
    filter(pathway == "GOBP_CILIUM_MOVEMENT") %>%
    pull(leadingEdge) %>%
    strsplit(";") %>%
    unlist() %>%
    unique()
}

# Pull leading-edge genes from each term's f-ranked  result
apoe_go <- read_csv("edgeR/tables/GSEA/ApoE/GSEA_ApoE_GO_BP.csv",
                           show_col_types = FALSE)
ex19_go <- read_csv("edgeR/tables/GSEA/Ex19/GSEA_Ex19_GO_BP.csv",
                           show_col_types = FALSE)
interaction_go <- read_csv("edgeR/tables/GSEA/interaction/GSEA_interaction_rank_F.csv",
                           show_col_types = FALSE)

apoe_genes <- get_leading_edge(apoe_go)
ex19_genes <- get_leading_edge(ex19_go)
interaction_genes <- get_leading_edge(interaction_go)

# Union of all three lists + Rec8 as a control
all_cilia_genes <- unique(c(apoe_genes, ex19_genes, interaction_genes, "Rec8", "Ctse"))

# Rank by FDR from the interaction DE result and keep top 20
# Rec8 is always included regardless of rank
top20_cilia_genes <- df_res_interaction %>%
  filter(gene_name %in% all_cilia_genes, gene_name != "Rec8") %>%
  arrange(FDR) %>%
  slice_head(n = 19) %>% # 19 + Rec8 = 20
  pull(gene_name) %>%
  c("Rec8", 
    "Ctse", top3_apoe, top3_ex19)

# Label each gene by which term(s) it came from
# Priority: Interaction > ApoE > Ex19
# Rec8 is labeled Ex19 
gene_source <- tibble(gene = top20_cilia_genes) %>%
  mutate(source = case_when(
    gene %in% top3_apoe ~ "ApoE",      
    gene %in% top3_ex19 ~ "Ex19",
    gene == "Rec8" ~ "Ex19",
    gene == "Ctse" ~ "Interaction",
    gene %in% interaction_genes ~ "Interaction",
    gene %in% apoe_genes ~ "ApoE",
    gene %in% ex19_genes ~ "Ex19"
  ))

# Save the final gene list
dir.create("edgeR/tables/GSEA/interaction", recursive = TRUE, showWarnings = FALSE)
write_csv(gene_source, "edgeR/tables/GSEA/interaction/cilia_top20_FDR_ranked.csv")

gene_source
```

Create the cilia heatmap with the above established cilia genes as well as the top three genes for apoe and exon19 found in 4.


```{r cilia-heatmap}
col_fun <- colorRamp2(c(-2, 0, 2), c("cornflowerblue", "white", "darkgoldenrod1"))

col_split <- factor(
  c(rep("+ex19/E3", 3), rep("\u0394ex19/E3", 3),
    rep("+ex19/E4", 3), rep("\u0394ex19/E4", 3)),
  levels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4")
)


bottom_ann <- HeatmapAnnotation(
  Group = anno_block(
    gp = gpar(fill = c("#3B5998", "#4BACC6", "#1F5C1F", "#92C46A")),
    labels    = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"),
    labels_gp = gpar(fontsize = 8, fontfamily = "Helvetica",col = "black")
  ),
  which = "column",
  show_annotation_name = FALSE
)

# Subset and scale normalized counts
top20_ids <- gene_names %>% filter(gene_name %in% c(top20_cilia_genes, top3_apoe, top3_ex19))

mat <- normalized_counts[rownames(normalized_counts) %in% top20_ids$gene_id, ]
mat_scaled <- t(scale(t(mat)))
rownames(mat_scaled) <- top20_ids$gene_name[
  match(rownames(mat_scaled), top20_ids$gene_id)]
row_split <- factor(
  gene_source$source[match(rownames(mat_scaled), gene_source$gene)],
  levels = c("Interaction", "Ex19", "ApoE")
)


# source term, aligned to actual matrix row order
source_order <- gene_source$source[match(rownames(mat_scaled), gene_source$gene)]
row_ann <- rowAnnotation(
  Term = source_order,
  show_annotation_name = FALSE,
  col  = list(Term = c(
    "Interaction" = "#E74C3C",
    "ApoE" = "#3498DB",
    "Ex19" = "#2ECC71"
  )),
  annotation_legend_param = list(Term = list(title = "Gene category", title_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
  labels_gp = gpar(fontsize = 8, fontfamily = "Helvetica")))
)
ht_cilia <- Heatmap(mat_scaled,
  col = col_fun,
  name = "Z-score",
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_row_slices = FALSE,
  show_row_dend = FALSE,
  cluster_columns = FALSE,
  column_split = col_split,
  row_split = row_split,
  row_gap = unit(3, "mm"),
  row_title_gp = gpar(fontsize = 9, fontfamily = "Helvetica"),
  bottom_annotation = bottom_ann,
  left_annotation = row_ann,
  row_names_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
  column_title = NULL,
  border = FALSE,
  row_dend_side = "left",
  heatmap_legend_param = list(
    title = "Z-Score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("-2", "-1", "0", "1", "2"),
    title_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
    labels_gp = gpar(fontsize = 7, fontfamily = "Helvetica")
  ))



svg("deliverables/Fig2_PanelC_cilia_heatmap.svg", width = 5, height = 5.5)
draw(ht_cilia)
dev.off()
write.csv(mat_scaled, "deliverables/cilia_heatmap_zscores.csv")
draw(ht_cilia)
cairo_pdf("deliverables/Fig2_PanelC_cilia_heatmap.pdf", width = 5, height = 5.5)
draw(ht_cilia)
dev.off()
```

Create a venn diagram that shows the genes used in the above heatmap and which lists they are in

```{r}
heatmap_genes <- rownames(mat_scaled)

venn_list <- list(
  "ApoE" = heatmap_genes[heatmap_genes %in% c(apoe_genes, top3_apoe, "Rec8")],
  "Ex19"  = heatmap_genes[heatmap_genes %in% c(ex19_genes, top3_ex19)],
  "ApoE x Ex19" = heatmap_genes[heatmap_genes %in% c(interaction_genes, "Ctse")]
)


venn_diagram <- ggVennDiagram(venn_list) + scale_fill_gradient(low="grey90",high = '#E74C3C')+ theme(text = element_text(family = "Arial"),
        plot.title = element_text(family = "Helvetica"))


ggsave("deliverables/Fig2_VennDiagram.svg",
       venn_diagram, width = 8, height = 3.5)

venn_diagram
```



```{r}
# Interaction summary figure 

# Re-test each term to get gene-level F statistics
f_apoe <- glmQLFTest(fit, coef = "apoeE4")$table$F
f_splicing <- glmQLFTest(fit, coef = "splicingdelta")$table$F
f_interaction <- glmQLFTest(fit, coef = "apoeE4:splicingdelta")$table$F

# Mean F ratio per term (same logic as Zhao Fig 1A)
mean_f_df <- tibble(
  term      = factor(
    c("ApoE", "Ex19", "ApoE \u00d7 Ex19", "Error"),
    levels = c("ApoE", "Ex19", "ApoE \u00d7 Ex19", "Error")
  ),
  mean_F = c(
    mean(f_apoe, na.rm = TRUE),
    mean(f_splicing, na.rm = TRUE),
    mean(f_interaction, na.rm = TRUE),
    1  
  ),
  fill_col = c("#4472C4", "#ED7D31", "#A9D18E", "#808080")
)

f_ratio_plot <- ggplot(mean_f_df, aes(x = term, y = mean_F, fill = fill_col)) +
  geom_col(color = "black", linewidth = 0.4, width = 0.65) +
  scale_fill_identity() +
  labs(
    x = NULL,
    y = "Mean F ratio",
    title = "Hippocampal transcriptome"
  ) +
  theme_bw() +
  theme(
    text             = element_text(family = "Arial"),
    panel.border     = element_blank(),
    axis.line.y      = element_line(linewidth = 0.5),
    axis.line.x      = element_line(linewidth = 0.5),
    axis.ticks       = element_line(linewidth = 0.4),
    axis.text.x      = element_text(size = 9, angle = 35, hjust = 1),
    axis.text.y      = element_text(size = 9),
    axis.title.y     = element_text(size = 10),
    plot.title       = element_text(size = 10, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor   = element_blank()
  )

f_ratio_plot

ggsave("deliverables/Fig2_PanelD_mean_F_ratio.svg",
       f_ratio_plot, width = 3, height = 3.5)
ggsave("deliverables/Fig2_PanelD_mean_F_ratio.pdf",
       f_ratio_plot, width = 3, height = 3.5, device = cairo_pdf)
```
Table for the mean values shows above 
```{r}
f_apoe <- glmQLFTest(fit, coef = "apoeE4")$table$F
f_splicing <- glmQLFTest(fit, coef = "splicingdelta")$table$F
f_interaction <- glmQLFTest(fit, coef = "apoeE4:splicingdelta")$table$F
mean(f_apoe, na.rm = TRUE)
mean(f_splicing, na.rm = TRUE)
mean(f_interaction, na.rm = TRUE)

 mean_f <-tibble(
  Term = c("ApoE", "Ex19", "ApoE × Ex19", "Error"),
  Mean_F = c(
    round(mean(f_apoe, na.rm = TRUE), 4),
    round(mean(f_splicing, na.rm = TRUE), 4),
    round(mean(f_interaction, na.rm = TRUE), 4),
    1
  )
)
 mean_f
```


```{r interaction linechart}

filtered_interaction %>% select(gene_name, logFC, FDR) %>% arrange(FDR)
df_res_interaction %>% filter(gene_name == "Serpina3n") %>% select(gene_name, logFC, FDR)
gene_id <- gene_names %>% filter(gene_name == "Stbd1") %>% pull(gene_id)
log2(normalized_counts[gene_id, ] + 1)

gene_id <- gene_names %>% filter(gene_name == "Rec8") %>% pull(gene_id)
log2(normalized_counts[gene_id, ] + 1)
```
Interaction profile plotes below, not currently needed

```{r eval = FALSE}
# ── Interaction profile plots (Panel D).

make_interaction_plot <- function(gene_name_str, title_str) {
  
  gene_id_val <- gene_names %>% filter(gene_name == gene_name_str) %>% pull(gene_id)
  
  plot_df <- tibble(
    sample = colnames(normalized_counts),
    expr = log2(normalized_counts[gene_id_val, ] + 1)
  ) %>%
    left_join(
      coldata %>% rownames_to_column("sample") %>% select(sample, apoe, splicing),
      by = "sample"
    ) %>%
    mutate(
      ex19 = ifelse(splicing == "plus", "+ex19", "\u0394ex19"),
      ex19 = factor(ex19, levels = c("+ex19", "\u0394ex19")),
      apoe = factor(apoe, levels = c("E3", "E4"))
    )
  
  # group means and SEM
  summary_df <- plot_df %>%
    group_by(apoe, ex19) %>%
    summarise(
      mean_expr = mean(expr),
      sem = sd(expr) / sqrt(n()),
      .groups = "drop"
    )
  
  ggplot() +
    # individual points
    geom_point(data = plot_df,
               aes(x = ex19, y = expr, color = apoe),
               position = position_dodge(width = 0.1),
               size = 1.5, alpha = 0.4) +
    # lines connecting means
    geom_line(data = summary_df,
              aes(x = ex19, y = mean_expr, color = apoe, group = apoe),
              linewidth = 0.8) +
    # mean points
    geom_point(data = summary_df,
               aes(x = ex19, y = mean_expr, color = apoe),
               size = 3) +
    # error bars
    geom_errorbar(data = summary_df,
                  aes(x = ex19, ymin = mean_expr - sem, ymax = mean_expr + sem, color = apoe),
                  width = 0.08, linewidth = 0.5) +
    scale_color_manual(values = c("E3" = "#3B5998", "E4" = "#1F5C1F"),
                       labels = c("APOE3", "APOE4")) +
    labs(x = "Apoer2 exon 19", y = "Log2 normalized expression", 
         color = NULL, title = title_str) +
    theme_classic() +
    theme(
      text = element_text(family = "Arial"),
      plot.title = element_text(size = 10, face = "bold"),
      axis.title = element_text(size = 9),
      axis.text = element_text(size = 8),
      legend.position = "bottom",
      legend.text = element_text(size = 8)
    )
}

p1 <- make_interaction_plot("Stbd1", "Stbd1")
p2 <- make_interaction_plot("")  # swap in cilia gene
p3 <- make_interaction_plot("Rec8", "Rec8")

panel_d <- p1 + p2 + p3 + plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")

panel_d

ggsave("deliverables/Fig2_PanelD_interaction_plots.svg",
       panel_d, width = 9, height = 3.5)
ggsave("deliverables/Fig2_PanelD_interaction_plots.pdf",
       panel_d, width = 9, height = 3.5, device = cairo_pdf)
```

6/30/26 Deliverables:

1. The mean values for each category

```{r}
f_apoe <- glmQLFTest(fit, coef = "apoeE4")$table$F
f_splicing <- glmQLFTest(fit, coef = "splicingdelta")$table$F
f_interaction <- glmQLFTest(fit, coef = "apoeE4:splicingdelta")$table$F
mean(f_apoe, na.rm = TRUE)
mean(f_splicing, na.rm = TRUE)
mean(f_interaction, na.rm = TRUE)

 mean_f <-tibble(
  Term = c("ApoE", "Ex19", "ApoE × Ex19", "Error"),
  Mean_F = c(
    round(mean(f_apoe, na.rm = TRUE), 4),
    round(mean(f_splicing, na.rm = TRUE), 4),
    round(mean(f_interaction, na.rm = TRUE), 4),
    1
  )
)
 mean_f
```

Generate a heatmap with the top 20 significant genes (FDR <0.1, |FC| > 1) from each category (APOE, ex19, Interaction)

```{r}
filter_for_heatmap <- function(df) {
  df %>% 
    dplyr::filter(FDR < 0.1 & abs(logFC) >1) %>%
    arrange(PValue) %>%
    head(40)
}
top20_apoe <- filter_for_heatmap(df_res_apoe_main_effect)
top20_ex19 <- filter_for_heatmap(df_res_splicing_main_effect)
top20_interaction <- filter_for_heatmap(df_res_interaction)

col_fun <- colorRamp2(c(-2, 0, 2), c("cornflowerblue", "white", "darkgoldenrod1"))

col_split <- factor(
  c(rep("+ex19/E3", 3), rep("\u0394ex19/E3", 3),
    rep("+ex19/E4", 3), rep("\u0394ex19/E4", 3)),
  levels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4")
)


bottom_ann <- HeatmapAnnotation(
  Group = anno_block(
    gp = gpar(fill = c("#3B5998", "#4BACC6", "#1F5C1F", "#92C46A")),
    labels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"),
    labels_gp = gpar(fontsize = 8, fontfamily = "Helvetica",col = "black")
  ),
  which = "column",
  show_annotation_name = FALSE
)

# Subset and scale normalized counts
all_top20 <- bind_rows(
  top20_apoe %>% mutate(source = "ApoE"),
  top20_ex19 %>% mutate(source = "Ex19"),
  top20_interaction %>% mutate(source = "Interaction")
)

mat <- normalized_counts[all_top20$gene_id, ]
mat_scaled <- t(scale(t(mat)))
rownames(mat_scaled) <- all_top20$gene_name



row_split <- factor(all_top20$source, levels = c("Interaction", "Ex19", "ApoE"))
source_order <- all_top20$source


# source term, aligned to actual matrix row order

row_ann <- rowAnnotation(
  Term = source_order,
  show_annotation_name = FALSE,
  col  = list(Term = c(
    "Interaction" = "#E74C3C",
    "ApoE" = "#3498DB",
    "Ex19" = "#2ECC71"
  )),
  annotation_legend_param = list(Term = list(title = "Gene category", title_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
  labels_gp = gpar(fontsize = 8, fontfamily = "Helvetica")))
)
ht_top20 <- Heatmap(mat_scaled,
  col = col_fun,
  name = "Z-score",
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_row_slices = FALSE,
  show_row_dend = FALSE,
  cluster_columns = FALSE,
  column_split = col_split,
  row_split = row_split,
  row_gap = unit(3, "mm"),
  row_title_gp = gpar(fontsize = 9, fontfamily = "Helvetica"),
  bottom_annotation = bottom_ann,
  left_annotation = row_ann,
  row_names_gp = gpar(fontsize = 5, fontfamily = "Helvetica"),
  column_title = NULL,
  border = FALSE,
  row_dend_side = "left",
  heatmap_legend_param = list(
    title = "Z-Score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("-2", "-1", "0", "1", "2"),
    title_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
    labels_gp = gpar(fontsize = 7, fontfamily = "Helvetica")
  ))

draw(ht_top20)

svg("deliverables/Fig3_top20_heatmap.svg", width = 5, height = 10)

draw(ht_top20)
dev.off()


```

Generate a venn diagram of all significant genes (FDR <0.1, |FC| > 1) from the categories and how they overlap 
```{r}
all_top20 %>% filter(gene_name %in% c("Rec8", "Ctse"))
```
```{r}
df_res_apoe_main_effect %>% filter(FDR < 0.1 & abs(logFC) > 1) %>% arrange(PValue) %>% mutate(rank = row_number()) %>% filter(gene_name %in% c("Rec8", "Ctse")) %>% select(gene_name, rank, PValue, FDR)

df_res_interaction %>% filter(FDR < 0.1 & abs(logFC) > 1) %>% arrange(PValue) %>% mutate(rank = row_number()) %>% filter(gene_name %in% c("Rec8", "Ctse")) %>% select(gene_name, rank, PValue, FDR)
```


```{r}
filter_for_venn <- function(df) {
  df %>% 
    dplyr::filter(FDR < 0.1 & abs(logFC) >1)
}
venn_apoe <- filter_for_venn(df_res_apoe_main_effect)
venn_ex19 <- filter_for_venn(df_res_splicing_main_effect)
venn_interaction <- filter_for_venn(df_res_interaction)

venn_genes_list <- list(
  "ApoE" = venn_apoe$gene_id,
  "Ex19"  = venn_ex19$gene_id,
  "ApoE x Ex19" = venn_interaction$gene_id
)

library(ggplot2)
venn_diagram <- ggVennDiagram(venn_genes_list) + scale_fill_gradient(low="grey90",high = '#E74C3C')+ theme(text = element_text(family = "Arial"),
        plot.title = element_text(family = "Helvetica"))+
  labs(subtitle = paste0("n = ", length(unique(c(venn_apoe$gene_id, venn_ex19$gene_id, venn_interaction$gene_id))), " genes (FDR < 0.1, |FC| > 1)"))

ggsave("deliverables/Fig2_AllCatVenn.svg",
       venn_diagram, width = 8, height = 3.5)

bind_rows(
  venn_apoe %>% select(gene_id, gene_name, logFC, FDR) %>% mutate(category = "ApoE"),
  venn_ex19 %>% select(gene_id, gene_name, logFC, FDR) %>% mutate(category = "Ex19"),
  venn_interaction %>% select(gene_id, gene_name, logFC, FDR) %>% mutate(category = "Interaction")
) %>% write_csv("deliverables/venn_significant_genes.csv")
venn_diagram

```
Generate heatmap with the genes that only fall under the 9+2 motile cilia and motile cilia CC terms. 


```{r}
gsea_f %>% 
  dplyr::filter(pathway %in% c(
    "GOCC_9PLUS2_MOTILE_CILIUM",
    "GOCC_MOTILE_CILIUM"
  )) %>%
  dplyr::select(pathway, padj, NES)



get_pathway_genes <- function(df) {
  df %>%
    filter(pathway %in% c("GOCC_9PLUS2_MOTILE_CILIUM", "GOCC_MOTILE_CILIUM")) %>%
    pull(leadingEdge) %>%
    strsplit(";") %>%
    unlist() %>%
    unique()
}

```

```{r}
table(all_top20$source)
```


```{r cilia-heatmap}
pathway_genes <- unique(c(cc_apoe_genes, cc_ex19_genes, cc_interaction_genes))
col_fun <- colorRamp2(c(-2, 0, 2), c("cornflowerblue", "white", "darkgoldenrod1"))

col_split <- factor(
  c(rep("+ex19/E3", 3), rep("\u0394ex19/E3", 3),
    rep("+ex19/E4", 3), rep("\u0394ex19/E4", 3)),
  levels = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4")
)


bottom_ann <- HeatmapAnnotation(
  Group = anno_block(
    gp = gpar(fill = c("#3B5998", "#4BACC6", "#1F5C1F", "#92C46A")),
    labels    = c("+ex19/E3", "\u0394ex19/E3", "+ex19/E4", "\u0394ex19/E4"),
    labels_gp = gpar(fontsize = 8, fontfamily = "Helvetica",col = "black")
  ),
  which = "column",
  show_annotation_name = FALSE
)


# Subset and scale normalized counts
pathway_gene_list <- gene_names %>% filter(gene_name %in% pathway_genes)
mat <- normalized_counts[rownames(normalized_counts) %in% pathway_gene_list$gene_id, ]
mat_scaled <- t(scale(t(mat)))
rownames(mat_scaled) <- pathway_gene_list$gene_name[
  match(rownames(mat_scaled), pathway_gene_list$gene_id)]

ht_cilia_paths <- Heatmap(mat_scaled,
  col = col_fun,
  name = "Z-score",
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  show_row_dend = FALSE,
  cluster_columns = FALSE,
  column_split = col_split,
  row_title_gp = gpar(fontsize = 9, fontfamily = "Helvetica"),
  bottom_annotation = bottom_ann,
  row_names_gp = gpar(fontsize = 4, fontfamily = "Helvetica"),
  column_title = paste0("Motile cilia CC leading-edge genes (n = ", length(pathway_genes), ")"),
  column_title_gp = gpar(fontsize = 10, fontfamily = "Helvetica", fontface = "bold"),

  border = FALSE,
  row_dend_side = "left",
  heatmap_legend_param = list(
    title = "Z-Score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("-2", "-1", "0", "1", "2"),
    title_gp = gpar(fontsize = 8, fontfamily = "Helvetica"),
    labels_gp = gpar(fontsize = 7, fontfamily = "Helvetica")
  ))
ht_cilia_paths
svg("deliverables/Fig3_cilia_CC_heatmap.svg", width = 5, height = 10)
draw(ht_cilia_paths)
dev.off()
write.csv(mat_scaled, "deliverables/cilia_CC_heatmap_zscores.csv")
```

```{r}
sig_pathway_genes <- c(
  df_res_apoe_main_effect %>% filter(gene_name %in% pathway_genes, FDR < 0.1) %>% pull(gene_name),
  df_res_splicing_main_effect %>% filter(gene_name %in% pathway_genes, FDR < 0.1) %>% pull(gene_name),
  df_res_interaction %>% filter(gene_name %in% pathway_genes, FDR < 0.1) %>% pull(gene_name)
) %>% unique()
length(sig_pathway_genes)
```

```{r}
length(unique(c(venn_apoe$gene_id, venn_ex19$gene_id, venn_interaction$gene_id)))
```


```{r}
apoe_cc <- read_csv("edgeR/tables/GSEA/ApoE/GSEA_ApoE_GO_CC.csv", show_col_types = FALSE)
ex19_cc <- read_csv("edgeR/tables/GSEA/Ex19/GSEA_Ex19_GO_CC.csv", show_col_types = FALSE)
cc_apoe_genes <- get_pathway_genes(apoe_cc)
cc_ex19_genes <- get_pathway_genes(ex19_cc)
cc_interaction_genes <- get_pathway_genes(gsea_f)
pathway_genes <- unique(c(cc_apoe_genes, cc_ex19_genes, cc_interaction_genes))
cilia_venn_list <- list(
  "ApoE" = pathway_genes[pathway_genes %in% cc_apoe_genes],
  "Ex19"  = pathway_genes[pathway_genes %in% cc_ex19_genes],
  "ApoE x Ex19" = pathway_genes[pathway_genes %in% cc_interaction_genes]
)

cilia_cc_venn <- ggVennDiagram(cilia_venn_list) +
  scale_fill_gradient(low = "grey90", high = "#E74C3C") +
  theme(text = element_text(family = "Arial")) +
  labs(subtitle = paste0("n = ", length(pathway_genes), " motile cilia CC leading-edge genes"))

cilia_cc_venn

ggsave("deliverables/Fig3_cilia_CC_venn.svg", cilia_cc_venn, width = 5, height = 3.5)
```

```{r}
length(intersect(cc_apoe_genes, cc_ex19_genes))
length(intersect(cc_apoe_genes, cc_interaction_genes))
length(intersect(cc_ex19_genes, cc_interaction_genes))
length(Reduce(intersect, list(cc_apoe_genes, cc_ex19_genes, cc_interaction_genes)))
```

